Forecasting the 2024 U.S. Presidential Election

### 2024-10-20 paper/paper.qmd
editted introduction working on data-sec

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "My subtitle if needed"
author: 
  - Tim Chen
  - Steven Li
  - Tommy Fu
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper presents a statistical model for forecasting the outcome of the 2024 U.S. Presidential Election. Using polling data from various sources, we develop multiple linear regression models to predict the percentage of support for the main candidates, Kamala Harris and Donald Trump, in different states. We aggregate these predictions to simulate the Electoral College vote and estimate the likelihood of either candidate winning. Our results suggest that neither candidate currently secures the required 270 electoral votes to win the presidency."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Example of installing a package after setting the mirror
install.packages("tidyverse")

install.packages("here")
library(tidyverse)
# Install janitor package if not installed
install.packages("janitor")

# Load the janitor package
library(janitor)

install.packages("caret")
install.packages("ROCR")
install.packages("rstanarm")
library(rstanarm)
install.packages("modelsummary")
library(modelsummary)

library(here)

@@

Forecasting elections has been a challenging task for political scientists and statisticians, especially given the complexity and volatility of electoral dynamics in the U.S. The 2024 U.S. Presidential Election is no exception, with polling data playing a critical role in shaping public and expert expectations. Pollsters use different methodologies and sample populations, which can significantly affect the predictions.

This paper aims to forecast the 2024 U.S. Presidential Election using multiple linear regression models based on polling data. We use various predictors such as sample size, pollster ratings, and state-level data to predict support for Kamala Harris and Donald Trump across the United States. We then aggregate these predictions to simulate the Electoral College outcome.

The rest of this paper is structured as follows: Section 2 describes the data used for this analysis. Section 3 outlines the models developed for each candidate and their corresponding results. Section 4 discusses the aggregated Electoral College predictions. Finally, Section 5 provides conclusions and suggestions for future research.

@@

The dataset used in this analysis contains poll data from various sources, including information about polling organizations, sample sizes, methodologies, and state-level data. It includes several key variables, such as:

Pollster rating: A score that represents the reliability of the polling organization.
Sample size: The number of respondents in each poll.
Support percentage: The percentage of respondents supporting each candidate.
State: The U.S. state where the poll was conducted or, in some cases, national polling data.
The data has been processed and cleaned to ensure that all variables are correctly aligned for regression analysis. The variables were transformed into numeric forms where necessary (e.g., percentage of support), and missing data was handled appropriately.

```{r}
#| echo: false
#| message: false
#| warning: false
# Load necessary libraries
library(dplyr)
library(tidyr)

@@

Our primary outcome variable is the percentage of support for Kamala Harris and Donald Trump in each poll. Figure 1 below shows the distribution of support for both candidates across the polls included in the dataset.

```{r}
#| echo: false
#| message: false
#| warning: false
# Summary statistics for percentage support (pct)
summary(data$pct)

# Plotting percentage support for different candidates
ggplot(data, aes(x = candidate_name, y = pct)) +
  geom_boxplot() +
  labs(title = "Support Distribution Across Candidates", x = "Candidate", y = "Percentage")

@@

# Modeling Support for the Candidates

## Kamala Harris
We begin by modeling the percentage of support for Kamala Harris using a linear regression model. The predictors include the sample size, pollster ratings (e.g., pollscore and transparency score), and state. This model aims to quantify how these variables influence her support across different polls.

@@

```{r}
# Plot the relationship for Kamala Harris
ggplot(harris_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Kamala Harris",
       x = "Sample Size",
       y = "Percentage")

@@

## Donald Trump
A similar linear regression model was applied to Donald Trump's data. The predictors remain the same, and the goal is to determine the factors that drive support for him.

@@

The results indicate that sample size has a weak but statistically significant negative effect on support for Trump. This suggests that larger polls tend to show slightly lower support for Trump, although the effect size is small.

Figure 3 shows the relationship between the sample size and support percentage for Donald Trump.
```{r}
# Plot the relationship for Kamala Harris
ggplot(trump_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Donald",
       x = "Sample Size",
       y = "Percentage")

@@

## Multiple Linear Regression Models
To better capture the complexity of voter support, we constructed multiple linear regression (MLR) models for both Kamala Harris and Donald Trump. These models incorporate several predictors, including pollster rating (represented by pollscore), transparency score, sample size, and state-level data. By accounting for these factors, the MLR models allow us to control for more variables that influence voter support and provide more accurate predictions.

### Kamala Harris Model
The MLR model for Kamala Harris takes into account multiple factors that may impact her support across different polls. This includes not just the sample size of the poll, but also how pollster reliability (numeric_grade), transparency of poll data (transparency_score), and state-level polling contribute to predicting voter preferences.

@@

### Donald Trump Model
Similarly, we constructed an MLR model for Donald Trump using the same set of predictors to assess the factors that influence his support across the country.
```{r}
#| echo: false
#| message: false
#| warning: false
# Load the saved MLR model for Donald Trump
mlr_trump_model <- readRDS(here::here(("models/mlr_trump_model.rds")))

@@

## Multicollinearity Check Using Variance Inflation Factor (VIF)
To ensure that the predictors used in both models do not exhibit multicollinearity, we checked the Variance Inflation Factor (VIF) for each predictor. High VIF values indicate multicollinearity, which can affect the stability and reliability of the model coefficients.

@@

## Stepwise Model Selection
To further optimize the MLR models, we performed stepwise model selection, which systematically adds or removes predictors to minimize the Akaike Information Criterion (AIC) and improve model fit.

@@

## Final Predictions for Electoral College Votes
Using the final models for both candidates, we predicted the percentage of support in each state and aggregated the results to simulate the Electoral College outcome. The predicted percentage of support for each candidate is used to determine the likely winner in each state.

@@

# Electoral College Prediction
To forecast the winner of the 2024 election, we aggregate the predicted percentages of support from our models for each state and calculate the Electoral College votes. The candidate with 270 or more Electoral College votes is predicted to win the election.

@@

# Discussion

## Key Findings
The multiple linear regression (MLR) models developed in this paper provide valuable insights into the factors that influence voter support for Kamala Harris and Donald Trump. Through the analysis of polling data, several key predictors were identified as significant, including state, pollster rating (pollscore), and transparency score.

@@

## Model Strengths
One of the main strengths of this analysis is the incorporation of multiple predictors that allow us to account for various factors influencing voter support. By considering state-level data, pollster ratings, and transparency scores, the models provide a more nuanced prediction of voter behavior compared to models that rely solely on national-level polling.

@@

## Limitations
Despite the strengths of the models, there are several limitations that should be addressed:

@@

# Appendix {-}

@@

# References

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "My subtitle if needed"
author: 
  - Tim Chen
  - Steven Li
  - Tommy Fu
thanks: "Code and data are available at: https://github.com/timchen0326/US_presidential_election_forecast_2024.git."
date: today
date-format: long
abstract: "This paper presents a statistical model for forecasting the outcome of the 2024 U.S. Presidential Election. Using polling data from various sources, we develop multiple linear regression models to predict the percentage of support for the main candidates, Kamala Harris and Donald Trump, in different states. We aggregate these predictions to simulate the Electoral College vote and estimate the likelihood of either candidate winning. Our results suggest that neither candidate currently secures the required 270 electoral votes to win the presidency."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

library(tidyverse)
library(ggplot2)
library(janitor)
library(rstanarm)
library(modelsummary)
library(here)
library(dplyr)
library(knitr)

@@

Forecasting elections has long been one of the most challenging tasks for political scientists, statisticians, and analysts alike. The complexity of predicting voter behavior in large, diverse electorates like that of the United States stems from numerous factors, including ever-shifting public opinion, rapidly changing political climates, and varying levels of voter engagement. The 2024 U.S. Presidential Election is no exception. With polarized voter bases, unpredictable external factors, and the growing influence of non-traditional media, the race remains highly uncertain, making reliable forecasting models more important—and difficult—than ever.

Central to any electoral forecast is the use of polling data. Polls provide snapshots of voter intentions at specific points in time, shaping both public perceptions and expert expectations. Yet, polling comes with its own set of challenges. Pollsters employ different methodologies—ranging from online surveys to phone interviews—each with its own inherent biases. Sample populations may vary in size and representativeness, potentially skewing results in favor of certain demographic groups. Moreover, polling accuracy can be further complicated by factors such as non-response bias, the changing nature of the electorate, and the rise of unconventional voting patterns, particularly among younger and minority voters.

In this paper, we attempt to forecast the outcome of the 2024 U.S. Presidential Election by developing statistical models based on polling data from a variety of sources. By leveraging multiple linear regression models, we aim to predict the percentage of support for the main candidates—Kamala Harris and Donald Trump—across different states. Our analysis incorporates key variables such as pollster reliability, sample size, and state-level demographics, all of which contribute to shaping the electoral landscape. Ultimately, we aggregate these state-level predictions to simulate the outcome of the Electoral College, providing insights into the likelihood of either candidate securing the 270 votes required to win the presidency.

The rest of this paper is structured as follows: @sec-data discusses the data used for this analysis, including key variables and sources. @sec-model outlines the models developed for each candidate and presents the corresponding results.
@sec-predict discusses the aggregated Electoral College predictions based on the model outputs. Finally, @sec-discuss provides conclusions and offers suggestions for future research.

@@

The dataset used in this analysis draws from a variety of polling sources, providing comprehensive information about polling organizations, sample sizes, methodologies, and state-level data. Several key variables are crucial to the analysis:

- Pollster rating: A numerical score reflecting the reliability and historical accuracy of each polling organization.
- Sample size: The number of respondents included in each poll, which influences the poll's margin of error.
- Support percentage: The percentage of respondents expressing support for each candidate.
- State: The U.S. state where the poll was conducted, or in some cases, national-level polling data.

To ensure accuracy and consistency, the data has been meticulously processed and cleaned. This involved aligning all variables for compatibility with regression analysis, converting relevant fields (e.g. support percentage) into numeric formats, and appropriately handling missing data to minimize potential biases. These reprocessing steps ensure the dataset is tidy and ready for model development.

@tbl-overview_data shows a subset of the dataset.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-overview_data
# Select key variables for the table
data_overview <- data %>%
  select(poll_id, pollster, pollscore, sample_size, pct, state, candidate_name, end_date)

# Show the first 5 rows of the dataset in a table
data_overview %>%
  head(5) %>%
  kable(caption = "Sample Overview of Selected Variables in the Polling Dataset")

@@

Our primary outcome variable is the percentage of support for Kamala Harris and Donald Trump in each poll.
@fig-support below shows the distribution of support for both candidates across the polls included in the dataset.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-support
#| fig-cap: Distribution of Support between Both Candidates

# Create a boxplot showing the distribution of support for each candidate
ggplot(data, aes(x = candidate_name, y = pct, fill = candidate_name)) +
  geom_boxplot() +
  labs(title = "Distribution of Support for Kamala Harris and Donald Trump",
       x = "Candidate",
       y = "Percentage of Support") +
  theme_minimal() +
  theme(legend.position = "none")

@@

# Modeling Support for the Candidates {#sec-model}

## Kamala Harris

We begin by modeling the percentage of support for Kamala Harris using a linear regression model.
The predictors include the sample size, pollster ratings (e.g., pollscore and transparency score), and state.
This model aims to quantify how these variables influence her support across different polls.

@@

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot the relationship for Kamala Harris
ggplot(harris_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Kamala Harris",
       x = "Sample Size",
       y = "Percentage")

@@

## Donald Trump

A similar linear regression model was applied to Donald Trump's data.
The predictors remain the same, and the goal is to determine the factors that drive support for him.

@@

The results in @figure-2 indicate that sample size has a weak but statistically significant negative effect on support for Trump.
This suggests that larger polls tend to show slightly lower support for Trump, although the effect size is small.

Figure 3 shows the relationship between the sample size and support percentage for Donald Trump.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: figure-2
# Plot the relationship for Kamala Harris
ggplot(trump_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Donald Trump",
       x = "Sample Size",
       y = "Percentage")

@@

## Multiple Linear Regression Models

To better capture the complexity of voter support, we constructed multiple linear regression (MLR) models for both Kamala Harris and Donald Trump.
These models incorporate several predictors, including pollster rating (represented by pollscore), transparency score, sample size, and state-level data.
By accounting for these factors, the MLR models allow us to control for more variables that influence voter support and provide more accurate predictions.

### Kamala Harris Model

The MLR model for Kamala Harris takes into account multiple factors that may impact her support across different polls.
This includes not just the sample size of the poll, but also how pollster reliability (numeric_grade), transparency of poll data (transparency_score), and state-level polling contribute to predicting voter preferences.

@@

### Donald Trump Model

Similarly, we constructed an MLR model for Donald Trump using the same set of predictors to assess the factors that influence his support across the country.

```{r}
#| echo: false
#| message: false
#| warning: false
# Load the saved MLR model for Donald Trump
mlr_trump_model <- readRDS(here::here(("models/mlr_trump_model.rds")))

@@

## Multicollinearity Check Using Variance Inflation Factor (VIF)

To ensure that the predictors used in both models do not exhibit multicollinearity, we checked the Variance Inflation Factor (VIF) for each predictor.
High VIF values indicate multicollinearity, which can affect the stability and reliability of the model coefficients.

@@

## Stepwise Model Selection

To further optimize the MLR models, we performed stepwise model selection, which systematically adds or removes predictors to minimize the Akaike Information Criterion (AIC) and improve model fit.

@@

## Final Predictions for Electoral College Votes

Using the final models for both candidates, we predicted the percentage of support in each state and aggregated the results to simulate the Electoral College outcome.
The predicted percentage of support for each candidate is used to determine the likely winner in each state.

@@

# Electoral College Prediction {#sec-predict}

To forecast the winner of the 2024 election, we aggregate the predicted percentages of support from our models for each state and calculate the Electoral College votes.
The candidate with 270 or more Electoral College votes is predicted to win the election.

@@

# Discussion {#sec-discuss}

## Key Findings

The multiple linear regression (MLR) models developed in this paper provide valuable insights into the factors that influence voter support for Kamala Harris and Donald Trump.
Through the analysis of polling data, several key predictors were identified as significant, including state, pollster rating (pollscore), and transparency score.

@@

## Model Strengths

One of the main strengths of this analysis is the incorporation of multiple predictors that allow us to account for various factors influencing voter support.
By considering state-level data, pollster ratings, and transparency scores, the models provide a more nuanced prediction of voter behavior compared to models that rely solely on national-level polling.

@@

## Limitations

Despite the strengths of the models, there are several limitations that should be addressed:

@@

# Appendix {.unnumbered}

@@

#References
